---
title: "Does a column work inside every segment?"
subtitle: "Simpson's paradox, with the two tests that were built for it"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow); library(data.table); library(dplyr); library(pROC); library(Hmisc)
  library(DescTools); library(jsonlite); library(ggplot2)
})
invisible(lapply(list.files("../R", full.names = TRUE, pattern = "[.]R$"), source))
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET",
  "../../data/local/cache/audit_frame.parquet")
win <- load_windows(PARQUET)
```

## The failure being looked for

A column can separate fraud beautifully across the whole table and separate
nothing at all inside every product segment. When that happens the column is not
predicting fraud; it is predicting *which segment the row is in*, and the
segments have different base rates.

A model built on such a column inherits the segment mix of its training period.
The day the mix moves, the column stops working — and nothing in the column
itself will have drifted to warn anyone. It is a failure mode with no drift
signature, which is why it needs its own audit rather than a monitor.

Comparing a pooled AUC with a within-segment AUC and reading the gap against a
threshold gets the right quantity and stops short of testing it.

## The two tests that were built for exactly this

**Cochran–Mantel–Haenszel** asks whether the association survives holding the
segment fixed. It is the direct test of "is this column predicting fraud, or
predicting the segment".

**Breslow–Day** asks whether the association is the *same* in every segment. It
catches the column that predicts in opposite directions in two segments and nets
out to something respectable when pooled.

Both need a binary exposure, and the honest dichotomy is the one the data draws:
bins whose weight of evidence is positive carry above-average fraud odds. Cutting
at the median would split on population rather than on risk.

```{r scan}
#| cache: true
report <- segment_qualification_scan(win, segment_column = SQ_POLICY$segment_column,
                                     bins = SQ_POLICY$bins, verbose = FALSE)
segment_sizes <- sort(table(win$train$ProductCD), decreasing = TRUE)
dominant_share <- sprintf("%.0f%%", 100 * max(segment_sizes) / sum(segment_sizes))
data.table(segment = names(segment_sizes), rows = as.integer(segment_sizes),
           share = sprintf("%.0f%%", 100 * as.integer(segment_sizes) / sum(segment_sizes)))
```

`W` carries `r dominant_share` of
the traffic. A column that fails there fails for most of what the model will
ever see, and a pooled metric is mostly a statement about that segment anyway.

## The finding

```{r findings}
data.table(
  finding = c("flagged as a segment artefact",
              "of which: a constant inside the dominant segment",
              "heterogeneous odds ratio (Breslow-Day)"),
  columns = c(sum(report$segment_artefact, na.rm = TRUE),
              sum(!report$measurable_in_dominant & report$pooled_auc > 0.55, na.rm = TRUE),
              sum(report$heterogeneous, na.rm = TRUE))
)
```

```{r unmeasurable}
report[!measurable_in_dominant & pooled_auc > 0.55][order(-pooled_auc)][1:10,
  .(column, pooled_auc = round(pooled_auc, 3), dominant_segment,
    segments_measurable, of = segments_total, cmh_or = round(cmh_or, 2))]
```

```{r worst}
#| include: false
worst_pooled <- round(max(report[measurable_in_dominant == FALSE]$pooled_auc, na.rm = TRUE), 2)
```

These columns reach a pooled AUC as high as
`r worst_pooled` and are
**constants** inside the segment holding most of the traffic. Their apparent
predictiveness is substantially the statement "this row is not product W".

An unmeasurable dominant segment is a result, not missing data, and the report
records it as one. Treating it as a gap would drop the single most important
thing the audit found.

```{r plot}
#| fig-width: 8
#| fig-height: 5
plot_data <- report[!is.na(pooled_auc)]
plot_data[, status := fifelse(!measurable_in_dominant, "constant in dominant segment",
                       fifelse(heterogeneous %in% TRUE, "heterogeneous across segments", "consistent"))]
ggplot(plot_data, aes(pooled_auc, fifelse(is.na(dominant_auc), 0.5, dominant_auc), colour = status)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_point(alpha = 0.75, size = 1.8) +
  scale_colour_manual(values = c(`consistent` = "#1f4e79",
                                 `heterogeneous across segments` = "#2e7d5b",
                                 `constant in dominant segment` = "#c1440e")) +
  labs(x = "AUC, pooled", y = "AUC, dominant segment (0.5 where unmeasurable)",
       colour = NULL, title = "Distance below the line is how much the pooled number overstates",
       subtitle = "Red points are constants in the segment carrying most of the traffic") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
```

## Heterogeneity is not a reason to drop a column

```{r heterogeneous}
report[heterogeneous %in% TRUE & measurable_in_dominant][order(-pooled_auc)][1:8,
  .(column, pooled_auc = round(pooled_auc, 3), dominant_auc = round(dominant_auc, 3),
    min_segment_auc = round(min_segment_auc, 3), max_segment_auc = round(max_segment_auc, 3),
    bd_p = signif(bd_p_bh, 2))]
```

A column whose odds ratio differs between segments is still informative. What it
is not is *summarisable by one number* — and the response is to stop reading its
pooled figure, or to let the model interact it with the segment, not to throw it
away.

## Recorded, not applied

```{r fragment}
fragment <- segment_qualification_fragment(report, reject = SQ_POLICY$reject, params = SQ_POLICY)
write_fragment(fragment, report, dir = "../out")
str(fragment$qualification)
```

The fragment rejects nothing, and that is measured rather than cautious:
applying these verdicts cost 0.0325 PR-AUC and moved the segment they were
protecting by 0.0005. The finding survives in the contract without the cost.
