---
title: "Does a feature still mean the same thing later?"
subtitle: "Time consistency, as a scorecard rather than as a model"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow); library(data.table); library(dplyr)
  library(pROC); library(Hmisc); library(jsonlite); library(ggplot2)
})
invisible(lapply(list.files("../R", full.names = TRUE, pattern = "[.]R$"), source))
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET",
  "../../data/local/cache/audit_frame.parquet")
```

## The question

A feature that separates fraud from non-fraud in January and separates it the
*other way* in June has not found a weak signal. It has found a pattern that
belongs to January. Training on it teaches the model a relationship that has
already expired, and the pooled metric will not complain, because the pooled
metric averages the two halves of the reversal.

The experiment is therefore always the same: measure early, measure late, and
compare. Everything below is about how to measure in a way that can be read.

## Why this is not a model

The obvious way to answer it is to fit a single-feature model per column and read
its AUC in both windows. That works, and it is a model used as a measuring
instrument, with three costs: the direction it learned cannot be recovered, a
rejection cannot be explained to anyone who has to approve the feature set, and
the verdict depends on hyperparameters that have nothing to do with the data.

A weight-of-evidence table answers the same question. It is also a learned,
possibly non-monotone mapping from values to a fraud ordering — that is exactly
what binning plus WoE *is* — but it prints as eleven rows. And the AUC of a
WoE-scored window is the Mann–Whitney statistic, so nothing is given up in
precision.

```{r load}
win <- load_windows(PARQUET)
c(train = nrow(win$train), holdout = nrow(win$holdout), features = length(win$features))
```

The windows are the first 17% and the last 17% of the time axis. The gap in the
middle is deliberate: it mimics the label-maturity lag you actually have at
deployment, where the most recent period has not finished being labelled yet.

```{r span}
axis <- data.table::as.data.table(arrow::read_parquet(PARQUET, col_select = "TransactionDT"))
w <- time_windows(axis$TransactionDT)
span <- function(i) diff(range(axis$TransactionDT[i])) / 86400
data.table(window = c("early", "late"),
           rows = c(length(w$train), length(w$holdout)),
           calendar_days = round(c(span(w$train), span(w$holdout)), 1))
```

**Equal rows, unequal calendar.** These are quantiles of the time value, so the
two windows hold the same number of transactions and cover 22.6 and 35.1 days
respectively — traffic falls over the period, so the later window watches for
1.56× as long. Part of what reads as "this feature changed" is therefore "the
later window watched for longer", and that confound is in every comparison
below.

It is left this way deliberately. Equal calendar spans would make the windows
hold unequal numbers of rows, and every statistic downstream would then have
different power on the two sides — a worse problem for a *test* than an unequal
span is for an estimate. What it rules out is reading any of this as a statement
about a fixed period; it is a statement about the first and last sixth of the
traffic.

## One column, in full

`V150` is the case worth looking at.

```{r v150}
y_tr <- as.integer(win$train$isFraud); y_ho <- as.integer(win$holdout$isFraud)
b <- fit_binning(win$train$V150)
early <- woe_table(assign_bins(b, win$train$V150), y_tr)
late  <- woe_table(assign_bins(b, win$holdout$V150), y_ho)

merge(early[, .(bin, n_early = n, woe_early = round(woe, 3))],
      late[,  .(bin, n_late  = n, woe_late  = round(woe, 3))],
      by = "bin", sort = FALSE)
```

Read it as a sentence. Early, higher values of `V150` meant sharply *lower*
fraud odds — the weight of evidence falls from −0.30 to −2.12 across the
populated range. Late, the two upper bins are empty: nothing in the later window
takes those values at all. What remains is the lowest bin, and its weight of
evidence has changed sign, from −0.30 to +0.49.

So the column did not merely weaken. Its populated range collapsed, and the odds
attached to what survived reversed. A model trained on the early window has
learned a rule about values that no longer occur, and applies the opposite of it
to the values that do.

That is the finding, and it took four rows to state.

## The scan

Fit the scorecard on the early window, apply it unchanged to the late one, and
read the AUC on both sides. Scoring the late window with weights learned *on the
late window* would ask a much weaker question — "is there any signal later" —
rather than "does the signal found earlier still hold".

```{r scan}
#| cache: true
report <- time_consistency_scan(win, opts = list(
  bins = TC_POLICY$bins, margin = TC_POLICY$margin,
  alpha = TC_POLICY$alpha, min_gini_drop = TC_POLICY$min_gini_drop), verbose = FALSE)
report[, .N, by = .(verdict)][order(-N)]
```

```{r plot}
#| fig-width: 8
#| fig-height: 5
plot_data <- report[!is.na(auc_train)]
p <- ggplot(plot_data, aes(auc_train, auc_holdout, colour = verdict)) +
  annotate("rect", xmin = 0.5, xmax = Inf, ymin = -Inf, ymax = 0.5,
           fill = "grey90", alpha = 0.6) +
  geom_hline(yintercept = 0.5, linewidth = 0.3) +
  geom_vline(xintercept = 0.5, linewidth = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_point(alpha = 0.75, size = 1.8) +
  scale_colour_manual(values = c(pass = "#1f4e79", inverted = "#c1440e",
                                 weak = "#8c8c8c", degenerate = "#2e7d5b")) +
  labs(x = "AUC, early window", y = "AUC, late window",
       colour = NULL,
       title = "The shaded quadrant is the finding",
       subtitle = "Signal early, reversed signal late. Points on the dashed line kept their meaning.") +
  theme_minimal(base_size = 12) + theme(panel.grid.minor = element_blank())
p
```

```{r save-plot}
#| echo: false
ggsave("../../docs/img/time-consistency-scatter.svg", plot = p,
       width = 8, height = 5, dpi = 300, bg = "white")
```

## Why a test and not a threshold

The verdict is a pair of tests: the later AUC must lie significantly past 0.5 by
a DeLong interval, the difference must survive a Benjamini–Hochberg correction
across all `r nrow(report)` columns, and the loss of discrimination must be
material.

The alternative is a fixed margin around 0.5 answering all three questions at
once. Applying one to the same AUCs shows what it costs:

```{r verdicts}
legacy <- fifelse(
  is.na(report$auc_train) | is.na(report$auc_holdout), "degenerate",
  fifelse(report$auc_train >= 0.52 & report$auc_holdout <= 0.48, "inverted",
  fifelse(abs(report$auc_train - 0.5) < 0.02 & abs(report$auc_holdout - 0.5) < 0.02,
          "weak", "pass")))
table(margin = legacy, tested = report$verdict)
```

Testing rejects *more*, and that is the point rather than a bug. At 100,000 rows
per window a DeLong interval is roughly 0.005 wide, so a margin of 0.02 was never
a significance threshold — it was four standard errors, and it passed every real
inversion smaller than itself.

Those rules are reconstructed here purely to show the difference. They are not a
policy this package holds and they appear nowhere in the report: a second
decision rule sitting next to the real one is a thing somebody eventually acts
on by mistake.

What stops inference alone from rejecting the whole table is the second key.
At this sample size almost any difference is significant, so significance is
necessary and not sufficient: the drop in Gini must also clear `min_gini_drop`.
A column is rejected only when the change is both real and large enough to care
about.

```{r multiplicity}
report[verdict == "inverted",
       .(feature, auc_train, auc_holdout, iv_train, woe_flip_mass,
         p_bh = signif(p_delong_bh, 3))][order(-woe_flip_mass)][1:10]
```

`woe_flip_mass` is the share of the later window sitting in bins whose weight of
evidence changed sign. It is the statistic that makes a rejection auditable:
when it is near 1, essentially the whole population moved into bins that now
mean the opposite of what they meant, which is what happened to `V150`.

## What leaves this notebook

```{r fragment}
fragment <- time_consistency_fragment(
  report,
  blocks = if (isTRUE(TC_POLICY$reject_by_block)) load_blocks("../../references/column-groups-v.json") else NULL,
  params = TC_POLICY)
write_fragment(fragment, report, dir = "../out")
str(fragment$qualification)
```

A fragment, and the table behind it. The fragment is a verdict plus the evidence
that produced it; nothing here decides admission, because admission needs every
check's fragment and the policy, and that merge happens once, in
`build-contract.qmd`.
