---
title: "Has the population moved?"
subtitle: "Stability, with the null distribution the conventional thresholds leave out"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow); library(data.table); library(dplyr); library(pROC); library(Hmisc)
  library(energy); library(twosamples); library(jsonlite); library(ggplot2)
})
invisible(lapply(list.files("../R", full.names = TRUE, pattern = "[.]R$"), source))
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET",
  "../../data/local/cache/audit_frame.parquet")
win <- load_windows(PARQUET)
```

## Two questions, routinely confused

**Did a column's marginal distribution move?** That is the population stability
index, and it is answered one column at a time.

**Are the two periods distinguishable at all?** That is a different question,
and a table can answer yes to it while every single margin answers no — because
the shift lives in the joint distribution.

Adversarial validation answers the second by training a model to tell the
periods apart and reading its AUC. That is a two-sample test with a model inside
it. Below is the same test without one: a p-value instead of an AUC, and a null
distribution instead of a convention.

## What PSI is when nothing has moved

The conventional bands — under 0.1 no shift, 0.1 to 0.25 moderate, over 0.25
significant — have a long service record and no sampling distribution. They are
quoted as if 0.25 meant something about evidence. It does not: PSI is a
statistic, it has a distribution under the null of "these two samples came from
the same population", and that distribution depends on the sample sizes and the
bin count.

So measure it. Pool the two windows, split them at random into the original
sizes, recompute. Anything that survives is noise, because the split was random
by construction.

```{r noise-floor}
#| cache: true
report <- distribution_shift_scan(win, B = DS_POLICY$permutations,
                                  bins = DS_POLICY$bins, verbose = FALSE)
floor_q99 <- median(report$psi_null_q99, na.rm = TRUE)
data.table(
  statistic = c("median 99th percentile of the null", "conventional 'significant' cut", "ratio"),
  value = c(round(floor_q99, 5), 0.25, round(0.25 / floor_q99))
)
```

At these sample sizes the index reaches about `r round(floor_q99, 5)` from
sampling noise alone. The threshold everyone uses sits **`r round(0.25/floor_q99)`
times higher**. It is not a significance level; it is a materiality convention
that has been mistaken for one.

The permutation is never actually run. PSI depends on the two windows only
through their bin counts, and the counts of a random split follow a multivariate
hypergeometric distribution exactly — so the null is drawn from that instead, at
one `rhyper` call per bin rather than a shuffle of 200,000 row indices. Eleven
minutes becomes fourteen seconds, and `tests/testthat/test-distribution-shift.R`
checks the two agree.

## Which means significance decides nothing here

```{r two-keys}
data.table(
  criterion = c("significant after BH", "material (PSI > 0.25)", "both"),
  columns = c(sum(report$p_perm_bh < 0.05, na.rm = TRUE),
              sum(report$psi > 0.25, na.rm = TRUE),
              sum(report$psi > 0.25 & report$p_perm_bh < 0.05, na.rm = TRUE))
)
```

At 100,000 rows per window essentially every column moves detectably. A test
that rejects that many has decided nothing, which is why significance is
necessary and not sufficient here. Materiality does the discriminating; the test
only rules out columns whose movement is indistinguishable from a coin flip.

## The finding a single index hides

```{r missingness}
report[!is.na(psi)][order(-psi)][1:8,
  .(column, psi = round(psi, 3), psi_values_only = round(psi_excl_null, 4),
    null_ref = round(null_share_ref, 3), null_cur = round(null_share_cur, 3),
    delta = round(null_share_delta, 3))]
```

`M7`, `M8` and `M9` have among the largest indices in the table, and almost none
of it is about their values. Each became roughly 45 percentage points *more
populated* between the windows; with the null bucket excluded their
distributions are essentially unchanged — `M7` moves by 0.0006.

That is a different finding from "this column drifted", and it calls for a
different response. A column whose values moved has become unreliable. A column
that started being *collected* has become more informative, and rejecting it for
instability throws away the improvement.

`id_31` is the opposite case: its pooled index is 1.38 and its value-only index
is higher still at 3.70. Browser version strings churn continuously, and there
the drift is entirely real.

```{r decomposition}
#| fig-width: 8
#| fig-height: 5
plot_data <- report[!is.na(psi) & !is.na(psi_excl_null)]
ggplot(plot_data, aes(pmax(psi, 1e-5), pmax(psi_excl_null, 1e-5))) +
  annotate("rect", xmin = 0.25, xmax = Inf, ymin = 1e-5, ymax = 0.25,
           fill = "grey90", alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_point(aes(colour = abs(null_share_delta) > 0.1), alpha = 0.7, size = 1.8) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c(`TRUE` = "#c1440e", `FALSE` = "#1f4e79"),
                      labels = c(`TRUE` = "missingness moved > 10pp", `FALSE` = "missingness stable")) +
  labs(x = "PSI, all buckets", y = "PSI, null bucket excluded", colour = NULL,
       title = "In the shaded band the shift is missingness, not values",
       subtitle = "Points on the dashed line moved for the reason the index suggests") +
  theme_minimal(base_size = 12) + theme(panel.grid.minor = element_blank())
```

## Are the periods distinguishable jointly?

The energy statistic is a weighted L2 distance between the two empirical
characteristic functions, and the test is a permutation test on it — no
distributional assumption, no model, and a p-value rather than an AUC to
interpret.

```{r energy}
#| cache: true
stable <- report[!is.na(psi)][order(psi)][1:12]$column
numeric_stable <- stable[vapply(stable, function(c) is.numeric(win$train[[c]]), logical(1))]
fit <- multivariate_shift(win, numeric_stable, n_sub = 1500L, R = 199L)
data.table(columns_tested = length(numeric_stable), n_per_window = fit$n_ref,
           energy_statistic = round(fit$statistic, 2), p_value = fit$p_value)
```

The twelve *most stable* columns by PSI were chosen deliberately: individually
none of them moved, and the question is whether they moved together. This is
exactly the case adversarial validation exists to catch, and it is caught here
without an adversary.

**The time column is never in the matrix.** With `TransactionDT` present any
discriminator separates the periods perfectly, because the periods are *defined*
by it, and the p-value that comes back says nothing about the features. This is
where the folklore that "adversarial AUC is 1 on this competition" partly comes
from — it is frequently an artefact of leaving the timestamp in. `load_windows`
excludes the time axis from the feature set for exactly this reason.

The asymmetry between the two questions is one-directional on this data, and
worth stating: columns clear the PSI threshold while remaining jointly
unremarkable, and none go the other way. A marginal shift large enough to matter
shows up in the joint test too; a joint shift does not require any single margin
to have moved.

It is subsampled, and not for convenience — the statistic is O(n²) in memory.

## Marginal tests, where they earn their place

Anderson–Darling rather than Kolmogorov–Smirnov as the headline: KS is most
sensitive near the median and least sensitive in the tails, and in fraud the
tails are the subject.

```{r marginal}
#| cache: true
top <- report[!is.na(psi)][order(-psi)][1:6]$column
top_numeric <- top[vapply(top, function(c) is.numeric(win$train[[c]]), logical(1))]
marginal_tests(win, top_numeric, n_sub = 4000L, nboot = 300L)
```

`cliffs_delta` is the effect size that belongs next to those p-values — a rank
statistic on the same scale as the Somers' D the association audit reports, so
"how far did it move" is answered in units the rest of the contract uses.

## What leaves this notebook

```{r fragment}
fragment <- distribution_shift_fragment(report, psi_threshold = DS_POLICY$psi_threshold,
                                        alpha = DS_POLICY$alpha, params = DS_POLICY)
write_fragment(fragment, report, dir = "../out")
str(fragment$qualification)
```
