---
title: "What the columns are made of"
subtitle: "Categorical dependence, latent dimension, and whether absence is random"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow); library(data.table); library(dplyr); library(pROC); library(Hmisc)
  library(DescTools); library(naniar); library(psych); library(jsonlite); library(ggplot2)
})
invisible(lapply(list.files("../R", full.names = TRUE, pattern = "[.]R$"), source))
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET",
  "../../data/local/cache/audit_frame.parquet")
win <- load_windows(PARQUET)
```

Three questions the per-column scans cannot reach, because each is about a
column's *construction* rather than its performance. None of them rejects
anything; all three change how the rejections elsewhere should be read.

## Does a column know more about the period than about fraud?

The drift audit asks whether a column moved. This asks something sharper: given
that it moved, is the movement all there is? Cramér's V puts both associations
on the same [0, 1] scale — dependence with the label, and dependence with which
window a row came from — so they can be compared directly.

The interval is bootstrapped, and the resampling is done on the contingency
table rather than on the rows: V depends on the two vectors only through their
table, and a bootstrap resample's table is a multinomial draw on the observed
cell proportions. One `rmultinom` replaces 500 passes over 200,000 rows.

Bergsma's bias correction is applied, because V is biased upward on sparse
tables and a fifty-level column against a binary outcome is sparse in its tail
however many rows there are.

```{r cramer}
#| cache: true
cat_report <- categorical_scan(win, R = 500L, verbose = FALSE)
cat_report[1:10, .(column, levels, v_label = round(v_label, 3),
                   v_window = round(v_window, 3),
                   window_ci = sprintf("[%.3f, %.3f]", v_window_lo, v_window_hi),
                   knows_period_better)]
```

```{r cramer-plot}
#| fig-width: 8
#| fig-height: 5
plot_data <- cat_report[!is.na(v_window)]
plot_data[, v_label_plot := fifelse(is.na(v_label), 0, v_label)]
p <- ggplot(plot_data, aes(v_label_plot, v_window, colour = knows_period_better)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_errorbar(aes(ymin = v_window_lo, ymax = v_window_hi), width = 0, alpha = 0.5) +
  geom_point(size = 2.2, alpha = 0.85) +
  scale_colour_manual(values = c(`TRUE` = "#c1440e", `FALSE` = "#1f4e79"),
                      labels = c(`TRUE` = "knows the period better",
                                 `FALSE` = "knows fraud at least as well")) +
  labs(x = "Cramér's V with the label", y = "Cramér's V with the window",
       colour = NULL, title = "Above the line, a column is a period marker",
       subtitle = "Bars are bootstrap percentile intervals on the window association") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
p
```

```{r save-plot-cramer}
#| echo: false
ggsave("../../docs/img/cramer-plot.svg", plot = p,
       width = 8, height = 5, bg = "white")
```

`M7`, `M8` and `M9` sit furthest above the line: dependence with the label
around 0.035, dependence with the window around 0.465. Whatever they are, they
are more than ten times better at saying *when* a row arrived than at saying
whether it was fraud.

This corroborates the drift audit through an entirely separate statistic. There
the finding was that their PSI is almost all missingness; here it is that what
they mainly encode is the period. Two methods, one conclusion, no shared
machinery.

## How many things is a block of V columns measuring?

The obvious response to a block of twenty correlated columns is to compress it.
Whether that is worth doing depends on the block's real dimension, and the usual
answers — eigenvalues above one, a scree plot read by eye — are conventions that
systematically overstate it on correlated data.

Horn's parallel analysis is the test: simulate data of the same shape with no
structure, and keep only the components whose eigenvalues beat what noise
produced.

```{r parallel}
#| cache: true
dims <- dimensionality_scan(win, load_blocks("../../references/column-groups-v.json"),
                            n_iter = 30L, verbose = FALSE)
dims[1:10, .(block, columns, components, compression = round(compression, 1),
             variance_first = round(variance_first_component, 3))]
```

`V143-V166` is eleven columns carrying one component, with 94% of the variance
in it. `V167-V183` is the same story. Across the pinned blocks the compression
ratio runs from about four to eleven.

That reframes the redundancy audit's result rather than contradicting it. Keeping
one representative per correlation group looked lossy; parallel analysis says the
blocks have far fewer dimensions than columns, so it is much less lossy than it
looked. It is also the argument *against* PCA here: if a block is essentially
one-dimensional, a representative column is nearly as good as a component and is
a column anyone can name.

## Are the M columns one thing measured nine ways?

Pearson correlation between two binary columns is bounded well below one by the
marginal split alone — two dichotomies that come from the same latent variable
can look weakly related simply because they were cut at different points. The
tetrachoric coefficient estimates the correlation of the latent continuous
variables underneath, which is the quantity the question is about.

```{r tetrachoric}
#| cache: true
m_cols <- grep("^M[0-9]", win$features, value = TRUE)
tc <- tetrachoric_matrix(win, m_cols)
data.table(binary_columns = tc$columns,
           mean_abs_tetrachoric = round(tc$mean_off_diagonal, 3))
```

At around 0.31 the answer is no. The M block is moderately intercorrelated and
not one latent thing measured repeatedly, so treating its members as
interchangeable would lose information — which is consistent with the redundancy
audit keeping most of them.

## Is a value missing at random?

The pipeline already treats missingness as informative: the binning gives nulls
their own bucket rather than imputing them. That is a modelling choice and until
now an untested one.

Little's test compares the means of the observed variables across missingness
patterns. A family whose columns all vanish together cannot be tested against
itself — in the pattern where the family is absent there is nothing left to take
a mean of — so each family is tested against columns that are present in
essentially every row. The question becomes: when this family is missing, do the
things we always see look different?

```{r mcar}
#| cache: true
miss <- missingness_scan(win, verbose = FALSE)
miss[1:10, .(size, null_share = round(null_share, 3), patterns, df = mcar_df,
             mcar_p = signif(mcar_p_bh, 3),
             indicator_auc = round(indicator_auc, 3), absence_is_signal, example)]
```

```{r mcar-summary}
data.table(
  finding = c("missingness families", "MCAR rejected after BH",
              "absence also separates fraud"),
  n = c(nrow(miss), sum(miss$mcar_p_bh < 0.05, na.rm = TRUE),
        sum(miss$absence_is_signal, na.rm = TRUE))
)
```

The family with no missingness at all comes back at p ≈ 0.99, which is the
control this table needed: the test is not simply rejecting everything.

For the rest, the answer is that absence is structured and it separates the
label. Imputing those columns would destroy information the model should have,
and the null bucket the binning already gives them is the right treatment —
now with a test behind it rather than an intuition.

What the test does **not** say is which kind of non-randomness this is. MAR and
MNAR are not distinguishable from observed data by any procedure, in R or
anywhere else. The claim available here is "not MCAR", and that is the claim
made.
