---
title: "What the fraud literature asks that the columns do not"
subtitle: "Five classical tests, none of which needs a model"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow); library(data.table); library(ggplot2)
  library(benford.analysis); library(poweRlaw); library(strucchange)
  library(circular); library(ineq)
})
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET",
  "../../data/local/cache/audit_frame.parquet")
d <- as.data.table(arrow::read_parquet(PARQUET,
  col_select = c("TransactionDT", "TransactionAmt", "isFraud", "card1", "addr1")))
```

Everything in the other notebooks asks the same question about every column:
does it carry signal, does it keep it, is it saying something the others do not.
This one asks the questions the fraud and forensic-statistics literature asks
about the *phenomenon*, which no per-column scan reaches.

Nothing here feeds the contract. These are report-only, and that is a real
distinction rather than a hedge: none of them names a column to drop. They
describe the thing being modelled, which is what a model card needs and what a
reviewer asks about first.

## Do the amounts obey Benford's law?

The canonical forensic-accounting test. In naturally occurring, multiplicatively
generated quantities the leading digit is 1 about 30% of the time and 9 about
4.6%; fabricated figures rarely reproduce that. It is the one test in this
repository an auditor will already know.

```{r benford}
#| fig-width: 8
#| fig-height: 4.5
fraud_amounts <- d[isFraud == 1 & TransactionAmt > 0]$TransactionAmt
bf <- benford(fraud_amounts, number.of.digits = 1)

digits <- data.table(
  digit = 1:9,
  observed = as.numeric(bf$bfd$data.dist),
  expected = as.numeric(bf$bfd$benford.dist)
)
ggplot(digits, aes(factor(digit))) +
  geom_col(aes(y = observed, fill = "observed"), width = 0.65) +
  geom_point(aes(y = expected, colour = "Benford"), size = 2.6) +
  geom_line(aes(y = expected, colour = "Benford", group = 1), linewidth = 0.6) +
  scale_fill_manual(values = c(observed = "#1f4e79")) +
  scale_colour_manual(values = c(Benford = "#c1440e")) +
  labs(x = "leading digit", y = "share", fill = NULL, colour = NULL,
       title = "Leading digit of fraudulent transaction amounts",
       subtitle = sprintf("n = %s", format(length(fraud_amounts), big.mark = ","))) +
  theme_minimal(base_size = 12) + theme(panel.grid.minor = element_blank())
```

```{r benford-stats}
data.table(
  statistic = c("mean absolute deviation", "Nigrini band", "chi-square p-value"),
  value = c(sprintf("%.5f", MAD(bf)),
            "0.012-0.015 = marginally acceptable conformity",
            format.pval(chisq(bf)$p.value))
)
```

And there is the theme of this whole audit again, in one line. The chi-square
test rejects conformity overwhelmingly. The effect size says the deviation is
marginal — right at the boundary of what Nigrini calls acceptable. At this
sample size the test detects a departure nobody would act on.

The correct reading is not "the amounts are fabricated". It is that the amounts
are *close* to Benford, close enough that the deviation is uninteresting, and
that a p-value on 20,000 observations was never going to say so.

## How heavy is the tail of the losses?

This is the one place where a distributional statistic touches a decision
already in the pipeline. `training/threshold.py` picks an operating point by
minimising an expected cost, and the cost of a missed fraud is its amount — so
the tail of the amount distribution is the tail of the loss distribution.

```{r powerlaw}
m <- conpl$new(fraud_amounts)
xmin_fit <- estimate_xmin(m)
m$setXmin(xmin_fit)
data.table(
  quantity = c("xmin (tail starts)", "alpha (tail index)", "observations in the tail",
               "finite variance?"),
  value = c(sprintf("%.2f", xmin_fit$xmin), sprintf("%.3f", xmin_fit$pars),
            format(sum(fraud_amounts >= xmin_fit$xmin), big.mark = ","),
            if (xmin_fit$pars > 3) "yes (alpha > 3)" else "no (alpha <= 3)")
)
```

An index near 3.7 means a tail heavy enough to matter and light enough to have
finite variance, so a cost expectation is well defined and the sample mean is a
usable estimator of it. Had it come out below 2, the expected cost the threshold
minimises would not exist and the whole calibration would need restating.

The honest caveat is the sample: `r format(sum(fraud_amounts >= xmin_fit$xmin), big.mark=",")`
observations above `xmin`. That is enough to reject a catastrophic tail and not
enough to pin the index tightly.

## When did the fraud rate change?

Comparing an early window with a late window, which is what the time-consistency
and drift audits do, presumes the change is between the windows. A structural
break test does not presume: it finds where the level changed, if it did.

```{r breaks}
#| fig-width: 8
#| fig-height: 4.5
daily <- d[, .(rate = mean(isFraud), n = .N), by = .(day = floor(TransactionDT / 86400))
           ][order(day)][n > 200]
bp <- breakpoints(rate ~ 1, data = daily, h = 0.1)
break_days <- daily$day[na.omit(bp$breakpoints)]

ggplot(daily, aes(day, rate)) +
  geom_line(colour = "#8c8c8c", linewidth = 0.4) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE, colour = "#1f4e79", linewidth = 0.8) +
  geom_vline(xintercept = break_days, colour = "#c1440e", linetype = "dashed", linewidth = 0.5) +
  labs(x = "day of the dataset", y = "fraud rate",
       title = "Daily fraud rate, with Bai-Perron breakpoints",
       subtitle = sprintf("breaks at days %s", paste(break_days, collapse = ", "))) +
  theme_minimal(base_size = 12) + theme(panel.grid.minor = element_blank())
```

Two breaks, and neither falls where the audit windows are cut. That is worth
knowing: the windows were chosen as the first and last 17% of the time axis, a
convention, and the data says the level actually moved at days
`r paste(break_days, collapse=" and ")`. A drift monitor pinned to the window
convention would report the change late.

## Is there a daily cycle?

Hour of day is a *circular* quantity: 23:00 and 01:00 are two hours apart, not
twenty-two. Every statistic that treats it as a number on a line — a mean, a
standard deviation, a regression coefficient — is wrong at the wrap-around, and
the wrap-around is exactly where night-time fraud lives.

Rayleigh's test is the right one: it tests uniformity on the circle against a
single-peaked alternative.

```{r circular}
#| fig-width: 7
#| fig-height: 5
hours <- (d[isFraud == 1]$TransactionDT / 3600) %% 24
angles <- circular(hours * 2 * pi / 24, units = "radians")
rt <- rayleigh.test(angles)
peak <- ((as.numeric(mean(angles)) %% (2 * pi)) * 24 / (2 * pi))

by_hour <- d[, .(rate = mean(isFraud), n = .N),
             by = .(hour = floor((TransactionDT / 3600) %% 24))][order(hour)]
ggplot(by_hour, aes(hour, rate)) +
  geom_col(fill = "#1f4e79", width = 0.9) +
  geom_hline(yintercept = mean(d$isFraud), linetype = "dashed", colour = "#c1440e") +
  coord_polar(start = 0) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  labs(x = NULL, y = "fraud rate",
       title = "Fraud rate by hour of the dataset clock",
       subtitle = sprintf("Rayleigh R = %.3f, p %s; concentration peaks at hour %.1f",
                          rt$statistic, format.pval(rt$p.value), peak)) +
  theme_minimal(base_size = 12)
```

The concentration is real and strong. What it is *not* is a statement about
local time: `TransactionDT` is seconds from an unpublished origin, so the peak is
at hour `r sprintf("%.1f", peak)` of the dataset's clock, not of anyone's day.
The cycle can be modelled; the hour cannot be named.

## How concentrated is the reconstructed client base?

The entity audit reconstructs a client from `card1` and `addr1` because the
dataset names no customer. Whether that reconstruction is *useful* is partly a
question of concentration: a key that gives almost every transaction its own
client explains nothing and memorises everything.

```{r concentration}
d[, uid := paste(card1, addr1)]
coverage <- d[, .N, by = uid]$N
data.table(
  statistic = c("clients reconstructed", "Gini of transactions per client",
                "Herfindahl-Hirschman index", "share of clients seen exactly once"),
  value = c(format(length(coverage), big.mark = ","),
            sprintf("%.3f", Gini(coverage)),
            sprintf("%.6f", sum((coverage / sum(coverage))^2)),
            sprintf("%.1f%%", 100 * mean(coverage == 1)))
)
```

A Gini of `r sprintf("%.3f", Gini(coverage))` says the transaction base is
heavily concentrated in a minority of reconstructed clients, and the
`r sprintf("%.0f%%", 100*mean(coverage==1))` seen exactly once are the reason the
validation split has to be by entity rather than by row: a client appearing once
in training and once in scoring is not a client the model learned, it is a client
the model memorised.

That is the number behind a choice the pipeline already makes, stated rather than
assumed.
