---
title: "What does this data look like?"
subtitle: "The descriptive pass, with intervals on every rate"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow); library(data.table); library(dplyr); library(pROC); library(Hmisc)
  library(ggplot2)
})
invisible(lapply(list.files("../R", full.names = TRUE, pattern = "[.]R$"), source))
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET", "../../data/local/cache/audit_frame.parquet")
d <- as.data.table(arrow::read_parquet(PARQUET))
```

The audits ask whether a column can be trusted. This asks what is in front of
them: how much fraud there is, where it sits, and which columns are what they
claim to be. Nothing here rejects anything.

Every rate carries a **Wilson interval** rather than a bare proportion. At a 3.5%
base rate, a segment with two hundred rows has a confidence interval about eight
percentage points wide, and a table of point estimates invites conclusions that
the sample sizes do not support.

```{r wilson}
#| include: false
wilson <- function(x, n, conf = 0.95) {
  z <- stats::qnorm(1 - (1 - conf) / 2)
  p <- x / n; denom <- 1 + z^2 / n
  centre <- (p + z^2 / (2 * n)) / denom
  half <- z * sqrt(p * (1 - p) / n + z^2 / (4 * n^2)) / denom
  list(lo = pmax(0, centre - half), hi = pmin(1, centre + half))
}
```

## How much fraud, and does the rate hold still?

```{r base-rate}
n <- nrow(d); k <- sum(d$isFraud)
ci <- wilson(k, n)
data.table(transactions = n, frauds = k,
           base_rate = sprintf("%.3f%%", 100 * k / n),
           ci_95 = sprintf("[%.3f%%, %.3f%%]", 100 * ci$lo, 100 * ci$hi),
           days = round(diff(range(d$TransactionDT)) / 86400, 1))
```

```{r drift}
#| fig-width: 8
#| fig-height: 4
daily <- d[, .(k = sum(isFraud), n = .N), by = .(day = floor(TransactionDT / 86400))][n > 200]
daily[, c("lo", "hi") := wilson(k, n)]
daily[, rate := k / n]

ggplot(daily, aes(day, rate)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#1f4e79", alpha = 0.18) +
  geom_line(colour = "#1f4e79", linewidth = 0.5) +
  geom_hline(yintercept = k / n, linetype = "dashed", colour = "#c1440e") +
  labs(x = "day of the dataset", y = "fraud rate",
       title = "Daily fraud rate with 95% Wilson bands",
       subtitle = "Dashed line is the pooled rate") +
  theme_minimal(base_size = 12) + theme(panel.grid.minor = element_blank())
```

The band is what stops a spike being read as an event: on a day with two thousand
transactions the interval is wide enough to contain most of the visible movement.
Where the level genuinely shifts is a different question, and it is answered with
a structural break test in
[what-the-fraud-literature-asks…](what-the-fraud-literature-asks-that-the-columns-do-not.qmd).

## Where the fraud is

```{r segments}
seg <- d[, .(n = .N, k = sum(isFraud)), by = ProductCD][order(-n)]
seg[, c("lo", "hi") := wilson(k, n)]
seg[, `:=`(rate = k / n, share_of_volume = n / sum(n))]
seg[, ratio_to_pooled := rate / (sum(k) / sum(n))]
seg[, .(ProductCD, n, share_of_volume = sprintf("%.1f%%", 100 * share_of_volume),
        fraud_rate = sprintf("%.2f%%", 100 * rate),
        ci = sprintf("[%.2f%%, %.2f%%]", 100 * lo, 100 * hi),
        vs_pooled = sprintf("%.1fx", ratio_to_pooled))]
```

The spread across products is the single largest source of fraud-rate variance in
the table, and the intervals do not overlap — this is a real segmentation, not a
sampling artefact. It is also why the promotion gate is judged on the *largest*
segment rather than on a pooled metric: `W` carries most of the volume at a
below-average rate, so a pooled number is mostly a statement about `W` while
looking like a statement about everything.

```{r device}
d[, has_device := !is.na(DeviceType)]
dev <- d[, .(n = .N, k = sum(isFraud)), by = has_device]
dev[, c("lo", "hi") := wilson(k, n)]
dev[, .(has_device, n, fraud_rate = sprintf("%.2f%%", 100 * k / n),
        ci = sprintf("[%.2f%%, %.2f%%]", 100 * lo, 100 * hi))]
```

Whether device information is *present at all* separates fraud by a factor of
several. That is the missingness-as-signal finding in its simplest form, and the
formal version — is the absence random, and does it carry the label — is in
[what-the-columns-are-made-of](what-the-columns-are-made-of.qmd).

## Does rarity predict fraud?

Frequency encoding replaces a category with how often it occurs. The premise is
that rarity itself is informative, and it is worth checking rather than assuming.

```{r rarity}
rarity_profile <- function(col) {
  x <- as.character(d[[col]])
  freq <- table(x)
  band <- cut(as.numeric(freq[x]), breaks = c(0, 10, 100, 1000, 10000, Inf),
              labels = c("<10", "10-99", "100-999", "1k-10k", ">10k"), right = FALSE)
  out <- data.table(band = band, y = d$isFraud)[!is.na(band),
          .(n = .N, k = sum(y)), keyby = band]
  out[, c("lo", "hi") := wilson(k, n)]
  out[, `:=`(column = col, rate = k / n, ratio = (k / n) / (sum(k) / sum(n)))]
  out[]
}
rbindlist(lapply(c("id_31", "DeviceInfo", "P_emaildomain"), rarity_profile))[
  , .(column, band, n, fraud_rate = sprintf("%.2f%%", 100 * rate),
      ci = sprintf("[%.2f%%, %.2f%%]", 100 * lo, 100 * hi),
      vs_column_base = sprintf("%.1fx", ratio))]
```

Monotone, and steeply so: the rarest browser versions run several times the
column's own base rate, and the relationship weakens smoothly as the level gets
more common. That is the argument for frequency encoding, and it is an argument
about *this* data rather than a technique borrowed on faith.

## What the amount says

```{r amount}
b <- fit_binning(d$TransactionAmt, bins = 10L)
tab <- woe_table(assign_bins(b, d$TransactionAmt), as.integer(d$isFraud))
tab[, .(bin, n, fraud = pos, share = round(share, 4), woe = round(woe, 3),
        iv_part = round(iv_part, 4))]
```

```{r amount-summary}
data.table(
  statistic = c("information value of the amount alone",
                "median amount, legitimate", "median amount, fraudulent",
                "mean amount, legitimate", "mean amount, fraudulent"),
  value = c(round(information_value(tab), 4),
            median(d[isFraud == 0]$TransactionAmt), median(d[isFraud == 1]$TransactionAmt),
            round(mean(d[isFraud == 0]$TransactionAmt), 2),
            round(mean(d[isFraud == 1]$TransactionAmt), 2))
)
```

The amount is a weak signal on its own — the means differ by around a tenth, and
the information value lands in the band conventionally read as weak. It is not
useless: the weight of evidence is monotone across the upper bins, so it carries
direction. It just cannot do much alone, which is worth knowing before anyone
builds a rule on it.

## Are the columns what they say they are?

```{r types}
profile <- rbindlist(lapply(setdiff(names(d), EXCLUDED_COLUMNS), function(cn) {
  x <- d[[cn]]
  data.table(column = cn,
             stored = class(x)[[1]],
             distinct = length(unique(x[!is.na(x)])),
             null_share = mean(is.na(x)))
}))
profile[, effectively := fifelse(distinct <= 2, "binary",
                          fifelse(distinct <= 20, "low-cardinality",
                          fifelse(stored %in% c("character", "factor"), "categorical", "continuous")))]
profile[, .N, by = .(stored, effectively)][order(-N)]
```

```{r types-mismatch}
# Stored as a number, behaving like a category: the case that matters, because
# every rank statistic in this package treats such a column as ordered and the
# ordering is an artefact of whatever integer the encoder happened to assign.
profile[stored %in% c("numeric", "integer") & distinct <= 20][order(distinct)][1:12,
  .(column, distinct, null_share = round(null_share, 3))]
```

These are the columns where the declared type and the behaviour disagree. The
audits handle them correctly by accident of design rather than by special case:
the binning falls back to one bin per distinct value when the quantile grid
collapses, which is exactly the right treatment for a discrete column, and the
weight-of-evidence path never assumes the ordering means anything.

Somers' D does assume it, which is why
[`association.R`](../R/association.R) returns `NA` for unordered categoricals
rather than coercing them — but for a *numeric* column with twelve levels it will
happily report an ordinal association that may be meaningless. Those columns are
read on their weight-of-evidence table, not on their D.
