---
title: "Which columns are saying the same thing twice?"
subtitle: "Correlation names a pair; redundancy names the one to drop"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow); library(data.table); library(dplyr); library(pROC); library(Hmisc)
  library(jsonlite); library(ggplot2)
})
invisible(lapply(list.files("../R", full.names = TRUE, pattern = "[.]R$"), source))
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET",
  "../../data/local/cache/audit_frame.parquet")
win <- load_windows(PARQUET)
tc <- fread("../out/tables/time_consistency.csv")
```

## Two claims under one word

"Redundant" is used for two different things and only one of them justifies
dropping a column.

**Correlated** is a pairwise, symmetric statement. It says two columns move
together. It does not say which one to remove, and on its own it is not even a
reason to remove either — two noisy measurements of the same quantity are worth
more than one.

**Predictable** is an asymmetric statement about a column and *all the others
taken together*. A column that a flexible fit reconstructs from its neighbours
contributes nothing they do not already carry, and that is a reason.

The audit asks both, and the gap between the answers is the finding.

## The hierarchy

```{r clusters}
#| cache: true
rho <- spearman_matrix(win, n_sub = 20000L)
clusters <- variable_clusters(rho, min_rho2 = 0.8)
data.table(
  columns = ncol(rho),
  groups = length(clusters$groups),
  singletons = sum(lengths(clusters$groups) == 1),
  largest_group = max(lengths(clusters$groups))
)
```

Distance is `1 - rho^2`, so what binds a group is *shared variance* rather than
the sign of the correlation: two columns moving exactly opposite each other
carry the same information and belong together. Average linkage rather than
complete, because complete linkage requires every pair in a group to clear the
threshold, and on a block of forty V columns that splits a family that is
plainly one family over a single loose member.

Spearman rather than Pearson because most of this table is zero-inflated counts
with heavy tails, where a Pearson coefficient is decided by a handful of extreme
rows.

## Where the threshold sits

A threshold nobody varied is a threshold nobody tested.

```{r sensitivity}
#| cache: true
#| fig-width: 8
#| fig-height: 4.5
grid <- rbindlist(lapply(seq(0.5, 0.95, by = 0.05), function(t) {
  cl <- variable_clusters(rho, min_rho2 = t)
  data.table(min_rho2 = t, groups = length(cl$groups),
             singletons = sum(lengths(cl$groups) == 1),
             largest = max(lengths(cl$groups)))
}))

p <- ggplot(grid, aes(min_rho2)) +
  geom_line(aes(y = groups, colour = "groups"), linewidth = 0.8) +
  geom_point(aes(y = groups, colour = "groups")) +
  geom_line(aes(y = largest, colour = "largest group"), linewidth = 0.8) +
  geom_point(aes(y = largest, colour = "largest group")) +
  geom_vline(xintercept = 0.8, linetype = "dashed", linewidth = 0.3) +
  scale_colour_manual(values = c(groups = "#1f4e79", `largest group` = "#c1440e")) +
  labs(x = expression(paste("minimum shared variance ", rho^2)), y = "columns",
       colour = NULL, title = "How the partition responds to the threshold",
       subtitle = "The dashed line is the policy setting") +
  theme_minimal(base_size = 12) + theme(panel.grid.minor = element_blank())
p
```

```{r save-plot-sensitivity}
#| echo: false
ggsave("../../docs/img/sensitivity.svg", plot = p,
       width = 8, height = 5, dpi = 300, bg = "white")
```

## Does the inherited V-block partition hold?

`references/column-groups-v.json` records a partition of the V columns taken
from published Kaggle work — human judgement read off correlation heatmaps,
pinned rather than re-derived. Whether it still describes *this* data is a
separate question from whether it was reasonable when it was written.

```{r blocks}
audit_pinned_blocks(rho, load_blocks("../../references/column-groups-v.json"))
```

Every block is tighter inside than out, the weakest by a factor of four. The
partition is inherited, and it holds.

## Correlated, and predictable

```{r redundancy}
#| cache: true
report <- redundancy_scan(win, tc_report = tc, min_rho2 = RD_POLICY$min_rho2,
                          r2 = RD_POLICY$redun_r2, n_sub = RD_POLICY$rank_subsample)
data.table(
  claim = c("non-representative (correlated with a group-mate)",
            "also reconstructable from the group",
            "correlated but carrying something of its own"),
  columns = c(sum(!report$is_representative),
              sum(!report$is_representative & isTRUE_na(report$predictable)),
              sum(!report$is_representative & !is.na(report$predictable) & !report$predictable))
)
```

That gap is the argument for asking twice. Dropping every non-representative
would remove `r sum(!report$is_representative)` columns; only
`r sum(!report$is_representative & isTRUE_na(report$predictable))` of them are
actually reconstructable from their group-mates. The rest correlate strongly and
still carry something nothing else does.

`Hmisc::redun` regresses each column on the rest using restricted cubic splines
and removes, one at a time, whichever is best predicted. Where the column is too
discrete to place three knots — 4,176 zeros and a thin tail is typical here —
the fit falls back to linear, and the report records which basis produced each
verdict rather than mixing them silently.

```{r basis}
report[, .N, by = .(basis = redun_note)][order(-N)]
```

## Choosing what stays

```{r representatives}
report[size >= 5][order(group, -is_representative)][1:12,
  .(group, column, size, representative = is_representative,
    null_rate = round(null_rate, 3), iv = round(iv, 3),
    predictable, r2 = round(redun_r2, 3))]
```

Least missing first, then the strongest information value among the survivors.
Missingness leads because a representative absent in 86% of rows represents its
group only where it is present, whatever its information value says.

## What leaves this notebook

```{r fragment}
fragment <- redundancy_fragment(report, params = RD_POLICY)
write_fragment(fragment, report, dir = "../out")
str(fragment$qualification)
```
