---
title: "Who is the customer, when the data does not say?"
subtitle: "Entity reconstruction, measured against what chance already gives you"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow); library(data.table); library(dplyr); library(pROC); library(Hmisc)
  library(ineq); library(ggplot2)
})
invisible(lapply(list.files("../R", full.names = TRUE, pattern = "[.]R$"), source))
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET", "../../data/local/cache/audit_frame.parquet")
win <- load_windows(PARQUET, features = c("card1", "card2", "card3", "addr1", "D1", "TransactionDT"))
```

## The problem

The dataset names no customer. Everything downstream needs one anyway: the
validation split has to be by entity or the model is scored on clients it
memorised, and the promotion gate judges the model on its cold-entity segment.

A customer can be reconstructed by pasting columns together — a card, an
address, and an anchor that separates two spells of the same card number. The
question is whether the reconstruction is *right*, and the test used for it is
label purity: when a label propagates within an entity, a correct reconstruction
produces label-homogeneous groups.

## Why purity alone says nothing

At a 3.5% base rate, a group of two rows is homogeneous 93% of the time by
chance. A reconstruction reported as "97% pure" is therefore not obviously doing
anything, and the number cannot be read without knowing what the same partition
would produce if the entity meant nothing at all.

So that is measured. The labels are permuted while the groups stay fixed, which
preserves both the base rate and the exact group-size distribution and destroys
only the association between entity and outcome.

```{r compare}
#| cache: true
candidates <- list(
  "card1"                = list(columns = "card1", anchor = NULL),
  "card1 + addr1"        = list(columns = c("card1", "addr1"), anchor = NULL),
  "card1 + addr1 + D1"   = list(columns = c("card1", "addr1"), anchor = "D1"),
  "card1 + card2 + addr1 + D1" = list(columns = c("card1", "card2", "addr1"), anchor = "D1")
)
keys <- compare_entity_keys(win, candidates, B = 200L, n_sub = 100000L)
keys[, .(candidate, rows_covered = round(rows_covered, 3), entities,
         singleton_share = round(singleton_share, 3),
         purity = round(pure_share, 4), null = round(null_mean, 4),
         lift = round(lift_over_null, 4), p = signif(p_value, 2))]
```

Read the last two columns against the one before them. Purity climbs steadily
with every component added — 0.909, 0.944, 0.985 — and the naive reading is that
the key keeps getting better.

The lift goes the other way: **0.089, 0.053, 0.040**. Each component added buys
less than the one before, because each one also shatters groups into singletons,
and the singleton share rises from 42% to 75%. A partition made mostly of
singletons is homogeneous by construction. The apparent improvement is the
measurement instrument responding to its own degradation.

```{r plot}
#| fig-width: 8
#| fig-height: 4.5
long <- rbind(
  keys[, .(candidate, quantity = "observed purity", value = pure_share)],
  keys[, .(candidate, quantity = "purity under shuffled labels", value = null_mean)]
)
long[, candidate := factor(candidate, levels = keys$candidate)]

ggplot(long, aes(candidate, value, fill = quantity)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(data = keys, aes(candidate, pure_share, fill = NULL,
                             label = sprintf("lift %.3f", lift_over_null)),
            vjust = -0.6, size = 3.4) +
  coord_cartesian(ylim = c(0.75, 1.02)) +
  scale_fill_manual(values = c("observed purity" = "#1f4e79",
                               "purity under shuffled labels" = "#8c8c8c")) +
  labs(x = NULL, y = "share of multi-row groups that are label-homogeneous",
       fill = NULL, title = "Almost all of the purity is the group sizes",
       subtitle = "The gap between the bars is the only part the reconstruction earned") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(), legend.position = "bottom",
        axis.text.x = element_text(angle = 12, hjust = 1))
```

## What this says about the key the pipeline uses

The pipeline reconstructs a client from `card1` and `addr1`, anchored on `D1`.
Two things in the table support that and one qualifies it.

Adding `card2` on top buys nothing — the lift is 0.0403 against 0.0405 — while
costing coverage. So the key is not under-specified.

The lift is highest for `card1` alone, which is *not* an argument for using
`card1`: it covers everything and leaves 42% singletons, so it groups
aggressively and its groups are large enough for purity to be informative. What
the numbers say is that the choice is a trade, and the pipeline's key sits at the
point where an added component stops paying.

Every candidate clears its null at p ≈ 0.01, the floor for 200 permutations. The
reconstruction is real. It is just far less impressive than 0.985 makes it sound,
and a validation split built on it inherits exactly that much confidence.

## Coverage and concentration

```{r coverage}
keys[, .(candidate, rows_covered = round(rows_covered, 3), entities,
         median_size, gini = round(gini, 3),
         median_span_days = round(median_span_days, 1))]
```

Coverage is what stops a key being judged only on the rows it happens to work
for: `card1 + addr1` reaches 91% of the table, and the 9% it misses are rows
where `addr1` is null — which is a population, not a random sample.

The Gini figure is the concentration of transactions across reconstructed
clients. A high one means a minority of entities carries most of the traffic,
which is what makes an entity-aware split necessary rather than fastidious.

## How much of the traffic has no history

The split and the gate both depend on this. A transaction whose entity was never
seen in training is one the model has no history for, and the share of those is
what the gate's cold-entity segment is measuring against.

```{r history}
frame <- win$train
uid <- entity_uid(frame, c("card1", "addr1"), anchor = "D1")
uid[is.na(frame$card1) | is.na(frame$addr1)] <- NA
seen_before <- duplicated(uid) & !is.na(uid)

data.table(
  situation = c("no reconstructable entity (a component is null)",
                "first transaction of its entity",
                "entity has prior history"),
  share = sprintf("%.1f%%", 100 * c(mean(is.na(uid)),
                                    mean(!is.na(uid) & !seen_before),
                                    mean(seen_before)))
)
```

Roughly half the table has no usable history at prediction time — either the key
cannot be built or the entity is being seen for the first time. Every aggregate
computed over an entity is therefore null-guarded, and the gate's cold-entity
check is not an edge case: it is most of the traffic.

## Does the chosen key match the one the pipeline builds?

The recommendation above is only worth something if the warehouse builds the same
thing. The key is constructed once, in the feature-engineering statement, and
carried into `model_input` as `client_uid`; this checks that the candidate this
analysis settles on reproduces it row for row — including which rows have no key
at all.

They are compared as *partitions* rather than as strings: the two use different
separators, so what has to match is which rows group together.

```{r verify}
check <- verify_against_pipeline_key(win$train, c("card1", "addr1"), anchor = "D1")
if (isTRUE(check$available)) {
  data.table(agreement = check$agreement,
             groups_here = check$groups_here, groups_pipeline = check$groups_pipeline,
             null_here = round(check$null_here, 4), null_pipeline = round(check$null_pipeline, 4))
} else {
  check$note
}
```

A disagreement here would not be a tie to break by preference. It would mean the
statement and this analysis describe different entities, and every purity figure
above is then about something other than what the model splits on.

## Where this decision goes

Nothing here writes a contract fragment. The verdict is a *choice of key*, and it
reaches the pipeline as the `entity` block of the contract — the components and
the anchor — where `training/` reads it to build the split and the gate reads it
to define its cold-entity segment.

The construction itself is one SQL expression, `features.CLIENT_UID_EXPRESSION`,
and Python reads the column it produces. Nothing rebuilds the identifier: an
identifier assembled in two places agrees only for as long as nobody edits
either. What lives here is the question of whether it is the right key, which is
a question about the data like every other one in this package — and the check
above is what ties the answer to the thing the pipeline actually runs.
