---
title: "Why the validation gap is not a mistake"
subtitle: "The embargo, the split, and what the daily rate says about both"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow)
  library(data.table)
  library(ggplot2)
})
invisible(lapply(list.files("../R", full.names = TRUE, pattern = "[.]R$"), source))
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET", "../../data/local/cache/audit_frame.parquet")
d <- data.table::as.data.table(
  arrow::read_parquet(PARQUET, col_select = c("TransactionDT", "isFraud"))
)
```

The pipeline splits the data four ways: **train**, an unassigned **gap**, **validation**,
and **test**. The gap is the part that looks like a bug and is not.

A chargeback arrives months after the transaction it belongs to. At deployment time,
the most recent period is therefore never finished being labelled: some frauds have not
yet been reported, so a model scored against that period is being scored against a label
that does not yet exist. Validating flush against training measures a situation nobody
ever has -- it reports optimism rather than performance. Leaving an unassigned strip
between train and validation is what forces the evaluation to happen on a period the
model never touched and that has had time to settle.

Everything else in the design follows from that one constraint.

## The split boundaries

The boundaries are **quantiles of the time axis**, not calendar dates. The dataset runs
a fixed number of rows; cutting at quantiles keeps the two audit windows the same size
in rows, which keeps every statistical test on the two sides at the same power. Calendar
cuts would not.

```{r windows}
w <- time_windows(d$TransactionDT,
  train   = AUDIT_WINDOWS$train,
  holdout = AUDIT_WINDOWS$holdout
)

# Full dataset split for context (train / gap / val / test)
q <- function(p) quantile_nearest(d$TransactionDT, p)
full_cuts <- sapply(c(0.75, 0.85, 0.90), q)
full_cuts_days <- as.numeric(full_cuts) / 86400

span <- function(idx) round(diff(range(d$TransactionDT[idx])) / 86400, 1)
data.table(
  segment = c(
    "train (0 -- 0.75)", "gap, unassigned (0.75 -- 0.85)",
    "validation (0.85 -- 0.90)", "test (0.90 -- 1.0)"
  ),
  rows = c(
    sum(d$TransactionDT <= q(0.75)),
    sum(d$TransactionDT > q(0.75) & d$TransactionDT <= q(0.85)),
    sum(d$TransactionDT > q(0.85) & d$TransactionDT <= q(0.90)),
    sum(d$TransactionDT > q(0.90))
  ),
  calendar_days = c(
    round((q(0.75) - min(d$TransactionDT)) / 86400, 1),
    round((q(0.85) - q(0.75)) / 86400, 1),
    round((q(0.90) - q(0.85)) / 86400, 1),
    round((max(d$TransactionDT) - q(0.90)) / 86400, 1)
  )
)
```

The audit windows used by the time-consistency and drift checks are the first 17%
and last 17% of the time axis (quantiles 0--0.17 and 0.83--1.0), which sit entirely
inside the training data and entirely outside it, respectively.

```{r audit-windows}
data.table(
  window = c("audit early (0 -- 0.17)", "audit late (0.83 -- 1.0)"),
  rows = c(length(w$train), length(w$holdout)),
  calendar_days = c(span(w$train), span(w$holdout)),
  note = c(
    "first sixth of the traffic",
    "last sixth, never seen in training"
  )
)
```

**Equal rows, unequal calendar.** Transaction volume falls over the period: the later
window covers 1.56x as many days as the early one for the same number of transactions.
That is a real confound -- part of what reads as "this feature changed" is "the later
window watched for longer". It is left this way because the alternative (equal calendar
spans) would give the two windows different row counts and therefore different statistical
power, which is a worse confound for a *test* than an unequal span is for an *estimate*.
The reproducibility check in the time-consistency notebook varies the windows and checks
whether verdicts survive.

## The daily rate

Wilson rather than the normal approximation: at a 3.5% base rate the normal interval
runs below zero on thin days, which is not a statement anyone should make about fraud.

The band around the line is the measurement's own resolution, and reading it is the
point of the figure. It widens where the day is thin, so a spike sitting inside a wide
band is a quiet day rather than an event -- the two are indistinguishable from the line
alone, which is why a bare daily rate is a misleading thing to plot. Behind the line the
four segments of the split are shaded, so every claim below can be checked against the
period it is about.

```{r rate}
#| fig-width: 8
#| fig-height: 5
s <- d[, .(n = .N, k = sum(isFraud == 1L)), by = .(day = as.integer(TransactionDT %/% 86400L))][order(day)]

z <- stats::qnorm(0.975)
s[, p := k / n]
s[, centre := (p + z^2 / (2 * n)) / (1 + z^2 / n)]
s[, half := (z / (1 + z^2 / n)) * sqrt(p * (1 - p) / n + z^2 / (4 * n^2))]
s[, `:=`(lo = pmax(0, centre - half), hi = pmin(1, centre + half))]

pooled <- sum(s$k) / sum(s$n)

cuts <- full_cuts_days
ymax <- max(s$hi)
bands <- data.frame(
  xmin = c(min(s$day), cuts[1], cuts[2], cuts[3]),
  xmax = c(cuts[1], cuts[2], cuts[3], max(s$day)),
  what = factor(c("train", "gap, unassigned", "val", "test"),
    levels = c("train", "gap, unassigned", "val", "test")
  )
)
bands$mid <- (bands$xmin + bands$xmax) / 2

p <- ggplot2::ggplot() +
  ggplot2::geom_rect(
    data = bands,
    ggplot2::aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = what),
    alpha = 0.5
  ) +
  ggplot2::scale_fill_manual(
    values = c(
      "train" = "#ffffff", "gap, unassigned" = "#f6e8d8",
      "val" = "#e8eef4", "test" = "#dbe4ed"
    ),
    guide = "none"
  ) +
  ggplot2::geom_vline(
    xintercept = cuts, colour = "#8c8c8c",
    linewidth = 0.35, linetype = "22"
  ) +
  ggplot2::geom_ribbon(
    data = s, ggplot2::aes(x = day, ymin = lo, ymax = hi),
    fill = "#1f4e79", alpha = 0.2
  ) +
  ggplot2::geom_hline(
    yintercept = pooled, colour = "#c1440e",
    linewidth = 0.5, linetype = "22"
  ) +
  ggplot2::geom_line(
    data = s, ggplot2::aes(x = day, y = p),
    colour = "#1f4e79", linewidth = 0.5
  ) +
  ggplot2::geom_text(
    data = bands, ggplot2::aes(x = mid, y = ymax * 1.06, label = what),
    colour = "#8c8c8c", size = 3.2, fontface = "bold", vjust = 0
  ) +
  ggplot2::annotate("text",
    x = min(s$day), y = pooled,
    label = paste0("pooled ", sprintf("%.1f%%", 100 * pooled), "  "),
    colour = "#c1440e", size = 3.2, hjust = 0, vjust = -0.6
  ) +
  ggplot2::scale_y_continuous(labels = function(x) paste0(round(100 * x, 1), "%")) +
  ggplot2::scale_x_continuous(expand = ggplot2::expansion(mult = c(0.005, 0.005))) +
  ggplot2::coord_cartesian(ylim = c(min(s$lo), ymax * 1.14)) +
  # Title and subtitle stay: the figure is embedded in the README, where it has to
  # say what it is without the surrounding notebook. The caption does not -- it was
  # six lines of argument rendered at 8.4pt, which is prose pretending to be a
  # graphic. That text now sits around this chunk, at full size and searchable.
  ggplot2::labs(
    title = "The fraud rate does not hold still, and some days barely measure it",
    subtitle = paste0(
      "Daily fraud rate across ", nrow(s),
      " days, shaded with a 95% Wilson interval. Pooled rate ",
      sprintf("%.2f%%", 100 * pooled), "."
    ),
    x = "Day on the transaction time axis", y = "Fraud rate"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(panel.grid.minor = ggplot2::element_blank())
p
```

The rate does not hold still. The first weeks sit visibly under the pooled line, which is
what a model trained on the whole period inherits: it is fitted partly on a regime that
the period it will be judged on has already left.

The Bai-Perron structural break test (in
[what-the-fraud-literature-asks...](what-the-fraud-literature-asks-that-the-columns-do-not.qmd))
finds two level shifts, neither of which falls where the audit windows are cut -- so the
windows are a convention the data does not endorse. A drift monitor pinned to those
boundaries would report the change late.

```{r save-svg}
#| echo: false
dir.create("../../docs/img", recursive = TRUE, showWarnings = FALSE)
ggplot2::ggsave("../../docs/img/daily-fraud-rate.svg",
  plot = p,
  width = 8, height = 5, dpi = 300, bg = "white"
)
```

## What the unassigned gap costs and what it buys

```{r gap-stats}
gap_rows <- sum(d$TransactionDT > q(0.75) & d$TransactionDT <= q(0.85))
gap_pct <- round(100 * gap_rows / nrow(d), 1)
gap_fraud <- d[TransactionDT > q(0.75) & TransactionDT <= q(0.85), sum(isFraud)]
data.table(
  quantity = c(
    "rows in the gap", "share of dataset",
    "fraud labels in the gap", "calendar span of the gap (days)"
  ),
  value = c(
    format(gap_rows, big.mark = ","),
    paste0(gap_pct, "%"),
    format(gap_fraud, big.mark = ","),
    as.character(round((q(0.85) - q(0.75)) / 86400, 1))
  )
)
```

The gap costs `r gap_pct`% of the labelled data, discarded rather than used for
training or evaluation. What it buys is a validation set that is actually comparable
to deployment: the model has not seen those rows, the labels are settled, and the
period is separated from training by enough time that a chargeback reported after
the split boundary is still a chargeback the validation set can see.

The test set (0.90 -- 1.0) is held back further still and is only used for the final
numbers in [MEASUREMENTS.md](../../docs/MEASUREMENTS.md).

## Why a fixed calendar gap would not work

A calendar embargo -- "skip the 30 days after training ends" -- looks equivalent and
is not. The dataset's transaction volume is uneven: a calendar cut leaves unequal
row counts in the two windows, and every test statistic downstream has different power
on the two sides. A quantile cut fixes the row count and lets the calendar span vary;
the commentary in the time-consistency audit is explicit about that confound and its
mitigation.

The policy setting is in `AUDIT_WINDOWS` in `R/policy.R`. A different value there
propagates automatically to every notebook that sources it -- there is one definition,
not two that have to be kept in sync.
