---
title: "The feature contract"
subtitle: "Every fragment, merged once"
format:
  html:
    toc: true
    code-fold: true
    df-print: paged
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(arrow); library(data.table); library(dplyr); library(pROC); library(Hmisc)
  library(jsonlite); library(ggplot2); library(DescTools)
})
invisible(lapply(list.files("R", full.names = TRUE, pattern = "[.]R$"), source))
PARQUET <- Sys.getenv("FRAUDAUDIT_PARQUET",
  "../data/local/cache/audit_frame.parquet")
```

This document computes nothing. Every verdict in it was decided by one of the
audit notebooks and written to `out/fragments/`; the only thing that happens
here is the merge, and the only decision is the order in which the checks get
credit for a column that more than one of them would have caught.

That order is policy. It does not change which columns are admitted — a column
rejected by two checks is rejected either way — but it decides what the funnel
looks like, so it is stated rather than left to whatever order the files were
listed in.

```{r fragments}
# PRECEDENCE comes from R/policy.R, and must match the one in
# `fraud_detection.cli.stamp_contract` -- the same order, stated twice, because
# the two sides render different things from it.
fragments <- read_fragments("out/fragments", order = PRECEDENCE)
data.table(
  check = names(fragments),
  tool = vapply(fragments, function(f) f$tool, character(1)),
  rejections = vapply(fragments, function(f) length(f$rejections), integer(1))
)
```

## What each check found

```{r qualification}
rbindlist(lapply(names(fragments), function(k) {
  q <- fragments[[k]]$qualification
  data.table(check = k, statistic = names(q),
             value = vapply(q, function(v) paste(format(unlist(v), trim = TRUE), collapse = ", "),
                            character(1)))
}))[]
```

## The merge

```{r contract}
win <- load_windows(PARQUET)
declared <- declare_columns(win,
  transaction_schema = "../schemas/bq_schema_train_transaction.json",
  identity_schema = "../schemas/bq_schema_train_identity.json")

policy <- lapply(fragments, function(f) f$params)
body <- build_contract_body(
  fragments, declared, policy,
  data_meta = list(table = "features.model_input",
                   rows = nrow(win$train) + nrow(win$holdout),
                   windows = win$windows))
str(body$summary, max.level = 1)
```

## Where columns are lost

```{r funnel}
#| fig-width: 8
#| fig-height: 4.5
cols <- rbindlist(lapply(body$columns, function(c) data.table(
  name = c$name, source = c$source, admitted = c$admitted,
  rejected_by = c$rejected_by)))

funnel <- rbindlist(list(
  data.table(stage = "declared", n = nrow(cols)),
  rbindlist(lapply(PRECEDENCE, function(k) data.table(
    stage = k, n = -sum(cols$rejected_by == k)))),
  data.table(stage = "admitted", n = sum(cols$admitted))
))
funnel[, stage := factor(stage, levels = funnel$stage)]

ggplot(funnel, aes(stage, abs(n), fill = n > 0)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = abs(n)), vjust = -0.4, size = 3.6) +
  scale_fill_manual(values = c(`TRUE` = "#1f4e79", `FALSE` = "#c1440e"), guide = "none") +
  labs(x = NULL, y = "columns",
       title = "Declared, then what each check removed, then what is left",
       subtitle = "Order is precedence: a column caught twice is credited to the earlier check") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(), axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r by-source}
cols[, .(declared = .N, admitted = sum(admitted),
         share = sprintf("%.0f%%", 100 * mean(admitted))), by = source][order(-declared)]
```

The `retrieved` row is the one with a deployment cost attached: every admitted
field there is a lookup the scoring path has to make before it can score.

## Is the admitted set better than a random one of the same size?

The question the old notebook answered with a model, asked without one. Total
information value is additive across columns and needs no fit, so the admitted
set can be compared directly with random sets drawn from the same declared pool.

```{r random-sets}
#| cache: true
tc <- fread("out/tables/time_consistency.csv")
iv <- tc[!is.na(iv_train), .(feature, iv_train)]
admitted_names <- cols[admitted == TRUE]$name
admitted_iv <- iv[feature %in% admitted_names]$iv_train
k <- length(admitted_iv)

set.seed(1)
null_draws <- replicate(2000, sum(sample(iv$iv_train, k)))
observed <- sum(admitted_iv)

data.table(
  quantity = c("admitted columns scored", "total IV of the admitted set",
               "mean total IV of a random set of the same size",
               "percentile of the admitted set", "one-sided p"),
  value = c(k, round(observed, 2), round(mean(null_draws), 2),
            sprintf("%.1f%%", 100 * mean(null_draws < observed)),
            signif((1 + sum(null_draws >= observed)) / 2001, 3))
)
```

```{r random-plot}
#| fig-width: 8
#| fig-height: 4
ggplot(data.table(iv = null_draws), aes(iv)) +
  geom_histogram(bins = 45, fill = "#8c8c8c") +
  geom_vline(xintercept = observed, colour = "#c1440e", linewidth = 0.9) +
  labs(x = "total information value", y = "random sets",
       title = "The admitted set against 2,000 random sets of the same size",
       subtitle = "Red line is the contract") +
  theme_minimal(base_size = 12) + theme(panel.grid.minor = element_blank())
```

The audits reject on *stability*, not on strength, so there was no guarantee the
surviving set would carry more signal than a random one — a set chosen for
carrying signal that expires would score well here and be worthless. That it
lands where it does is evidence the checks are not simply removing the strongest
columns.

## Is the admitted set stable across nearby policy settings?

A contract that flips when a threshold moves by a hundredth is a contract that
records the threshold rather than the data.

```{r stability}
#| cache: true
ds <- fread("out/tables/distribution_shift.csv")
tc_rejected <- tc[verdict == "inverted"]$feature

grid <- rbindlist(lapply(c(0.10, 0.15, 0.20, 0.25, 0.30, 0.40, 0.50), function(th) {
  shifted <- ds[!is.na(psi) & psi > th & p_perm_bh < 0.05]$column
  admitted_here <- setdiff(cols$name, union(tc_rejected, shifted))
  data.table(psi_threshold = th, admitted = length(admitted_here),
             jaccard_vs_policy = length(intersect(admitted_here, admitted_names)) /
               length(union(admitted_here, admitted_names)))
}))
grid[]
```

```{r stability-plot}
#| fig-width: 8
#| fig-height: 4
ggplot(grid, aes(psi_threshold, jaccard_vs_policy)) +
  geom_line(colour = "#1f4e79", linewidth = 0.8) +
  geom_point(size = 2.2, colour = "#1f4e79") +
  geom_vline(xintercept = 0.25, linetype = "dashed", linewidth = 0.3) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "PSI threshold", y = "Jaccard overlap with the policy setting",
       title = "How much of the admitted set survives moving the drift threshold",
       subtitle = "Dashed line is the setting the contract was stamped at") +
  theme_minimal(base_size = 12) + theme(panel.grid.minor = element_blank())
```

The curve is flat near the policy setting and falls away at the ends, which is
what a threshold sitting on a plateau rather than on a cliff looks like. The
number to quote is not the threshold; it is the width of the region over which
the answer does not change.

## What leaves this document

```{r write}
path <- write_declaration(
  declared,
  data_meta = list(table = "features.model_input",
                   rows = nrow(win$train) + nrow(win$holdout),
                   windows = win$windows),
  path = "out/declaration.json")
cat(readLines(path, n = 18), sep = "\n")
```

The declaration, and nothing else. The merged contract above is this document's
output — something for a person to read — and writing it to disk as well would
put a near-copy of `references/feature-contract.json` in the repository, which is
a second copy of a decided thing and therefore a thing that can disagree with the
first.

What stamping needs is the column list with its sources and dtypes; every verdict
is already in the fragments. It carries no fingerprint, because
`FeatureContract.from_dict` rejects a file whose stored hash disagrees with its
contents and that detector should not depend on two JSON serialisers agreeing
forever.

```bash
uv run stamp-contract
```

Everything downstream of that command — training, the promotion gate, batch
scoring — is Python's, and reads only the stamped file.
